Do larger watersheds respond different to forest cover change? Re-analysing a global data set.

### 2021-11-28 paper/Forest_and_Water/Forest_and_Water.tex
further changes to the paper

There has been an long and on-going discussion in the hydrologcal
literature around the impact of forests on streamflow (Andréassian,
2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al.,
2005; Zhang et al., 2017). The historic work highlights a general
consensus that if forest areas increase, streamflow decreases and
vice-versa. The most dramatic result in relation to this, is Figure 5 in
Zhang et al. (2011) indicating (for Australian watersheds) a 100\%
decrease in stream flow for watersheds with 100\% forest cover. However,
on the other end of the spectrum, in a series of French watersheds
(Cosandey et al., 2005), there was no change in streamflow
characteristics in 2 of the three watersheds studied in relation to
deforestation.

@@

Following Zhang et al. (2017), the first step is to use a linear
regression to investigate the percent change in flow as a result in the
percent change forestry and modulated by the direction of the change,
either an increase in forest cover, or decrease in forest cover.

@@

While the overall variance explained in this model is not high at 0.19,
it clearly supports the hypothesized relationship between the change in
forest cover and the change in flow. The model suggests that for every
1\% change in forest cover, on the average, the flow changes 0.5\%.
However the change in flow is different for forest cover decreases
compared to forest cover increases. In fact, forest cover increases
decrease flow by 31\% less than a similar decrease in forest cover
causes flow to increase. So roughly speaking, a 1\% forest cover
increase on the average decreases flow by \((1 - 0.31)*0.5\%\), while a
the percentage forest cover decrease will increase flow by 0.5\%.

It is however clear from the lack of explaining power, that there could
be confounding factors, as alluded to in the methods. The obvious ones
being watershed dryness and area (following Zhang et al. (2017)).

\begin{longtable}[]{@{}ccccc@{}}
\toprule
\begin{minipage}[b]{0.31\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{(Intercept)}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
18.94\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
7.78\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
2.43\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.02\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{DeltaF\_perc\_pos}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.5\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.09\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
5.66\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Forest\_Signincrease}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-31.54\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
5.9\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.35\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Area\_km2}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.3\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.77\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Pa\_mm}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.75\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.08\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
\end{figure}

As indicated, a second hypothesis relates to whether there is a strong
spatial global gradient as captured by latitude and longitude. As the
global map (@ref(fig:global\_map)) shows, the distribution of case study
watersheds covers multiple continents and shows some distinct clustering
in parts of the world. Of interest is whether the spatial clustering
also indicates a difference in response to forest cover change.

\begin{longtable}[]{@{}ccccc@{}}
\toprule
\begin{minipage}[b]{0.31\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{(Intercept)}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
29.99\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
8.65\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
3.47\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{DeltaF\_perc\_pos}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.47\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.09\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
5.4\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Forest\_Signincrease}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-37.11\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
6.09\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-6.09\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Area\_km2}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.59\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.55\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Pa\_mm}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.17\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.03\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Latitude}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.29\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.11\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.74\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.01\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Longitude}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.01\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.03\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
0.28\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.78\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

This linear model shows that there is a significant gradient in the
Latitude and with annual average rainfall, with watersheds closer to the
equator having lower changes in the runoff compare to watersheds further
away from the equator. This suggests an influence of radiation, which
will be tested next. In addition, the model suggests an influence of the
annual average rainfall, with wetter watersheds having slightly lower
changes in runoff. The total explaining power of the model is still low
at 0.22 suggesting further confounding factors currently not included in
the model.

There is no relationship with Longitude, suggesting that the different
continents do not show a trend in the East-West direction.

@@

The previous analysis suggests an influence of radiation on
evapotranspiration, and most likely related to the dryness index, as
also indicated in Zhang et al. (2017). Increased evapotranspiration
could lead to drier watersheds, unless balanced by rainfall (such as
possibly in the tropics). This model introduces the dryness index as a
linear variable and drops the annual average precipitation as a
variable.

@@

However, the result also indicates possible issues with the data, some
of the Dryness values are very large (\textgreater{} 4) and these values
have high leverage in the data. These watershed are listed in Table XX:

>>>>

There has been an long and on-going discussion in the hydrologcal
literature around the impact of forests on streamflow (Andréassian,
2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al.,
2005; Zhang et al., 2017). The historic work highlights a general
consensus that if forest areas increase, streamflow decreases and
vice-versa. The most dramatic result in relation to this, is Figure 5 in
Zhang et al. (2011) indicating (for Australian watersheds) a 100\%
decrease in streamflow for watersheds with 100\% forest cover. However,
on the other end of the spectrum, in a series of French watersheds
(Cosandey et al., 2005), there was no change in streamflow
characteristics in 2 of the three watersheds studied in relation to
deforestation.

@@

Following Zhang et al. (2017), the first step is to use a linear
regression to investigate the percent change in flow as a result in the
percent change forestry and modulated by the direction of the change,
either an increase in forest cover, or decrease in forest cover:

\[\tag{3}
\begin{aligned}
\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
\end{aligned}\]

@@

While the overall variance explained in this model is not high with an
adjusted \emph{r\textsuperscript{2}} of 0.19, it clearly supports the
hypothesized relationship between the change in forest cover and the
change in flow. The model suggests that for every 1\% change in forest
cover, on the average, the flow changes 0.5\%. However the change in
flow is different for forest cover decreases compared to forest cover
increases. In fact, forest cover increases decrease flow by 31\% less
than a similar decrease in forest cover causes flow to increase. So
roughly speaking, a 1\% forest cover increase on the average decreases
flow by \((1 - 0.31)*0.5\%\), while a the percentage forest cover
decrease will increase flow by 0.5\%.

It is however it is clear from the lack of explaining power, that there
could be confounding factors, as alluded to in the methods. The obvious
ones being watershed dryness and area (following Zhang et al. (2017)):

\[\tag{4}
\begin{aligned}
\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + \varepsilon
\end{aligned}\] Where \(Pa_mm\) is the annual average rainfall in mm.

\begin{longtable}[]{@{}ccccc@{}}
\caption{Summary of the second model, taking into account the annual
rainfall and the area of the watershed}\tabularnewline
\toprule
\begin{minipage}[b]{0.31\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{0.31\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{(Intercept)}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
18.94\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
7.78\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
2.43\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.02\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{DeltaF\_perc\_pos}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.5\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.09\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
5.66\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Forest\_Signincrease}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-31.54\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
5.9\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.35\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Area\_km2}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.3\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.77\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Pa\_mm}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.75\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.08\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included watersheds across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
\end{figure}

As indicated, a further hypothesis relates to whether there is a strong
spatial global gradient as captured by latitude and longitude. As the
global map (@ref(fig:global\_map)) shows, the distribution of case study
watersheds covers multiple continents and shows some distinct clustering
in parts of the world. Of interest is whether the spatial clustering
also indicates a difference in response to forest cover change:

\[\tag{5}
\begin{aligned}
\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + Latitude + Longitude + \varepsilon
\end{aligned}\]

\begin{longtable}[]{@{}ccccc@{}}
\caption{Results of the model including Latitude and
Longitude}\tabularnewline
\toprule
\begin{minipage}[b]{0.31\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{0.31\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{(Intercept)}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
29.99\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
8.65\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
3.47\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{DeltaF\_perc\_pos}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.47\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.09\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
5.4\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Forest\_Signincrease}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-37.11\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
6.09\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-6.09\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Area\_km2}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.59\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.55\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Pa\_mm}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.17\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.03\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Latitude}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.29\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.11\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.74\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.01\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.31\columnwidth}\centering
\textbf{Longitude}\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.01\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
0.03\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
0.28\strut
\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
0.78\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

This linear model shows that there is a significant gradient in the
Latitude and with annual average rainfall, with watersheds closer to the
equator having lower changes in the runoff compare to watersheds further
away from the equator. This suggests an influence of radiation, which
will be tested next. In addition, the model suggests an influence of the
annual average rainfall, with wetter watersheds having slightly lower
changes in runoff. The total explaining power of the model is still low
with an adjusted \emph{r\textsuperscript{2}} of 0.22 suggesting further
factors that are currently not included in the model.

There is no relationship with Longitude, suggesting that the watersheds
across different continents do not show an East-West direction trend.

@@

Latitude might be indicating an influence of radiation on
evapotranspiration, and most likely related to the dryness index, as
also indicated in Zhang et al. (2017). Increased evapotranspiration
could lead to drier watersheds, unless balanced by rainfall (such as
possibly in the tropics). The North-South trend in dryness would be
related to Latitude. This model introduces the dryness index as a linear
variable and drops the annual average precipitation as a variable, as
dryness is calculated from.

@@

However, the result also indicates possible issues with the data, some
of the Dryness values are very large (\textgreater{} 4) and these values
have high leverage in the data. These watersheds are listed in Table XX:
